---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Thomas, James"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)
# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS,mydata$RINGS)
```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: (The Variable types in mydata are numerical, integer and factor. The variables WHOLE, SHUCK, and VOLUME have max values more prominent than the mean and median. These may be potential outliers that lead to right-skewed distributions.)***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
library(RColorBrewer)
part_1b <- table(mydata$SEX,mydata$CLASS)
addmargins(part_1b, FUN = sum)
barplot(part_1b, main = "CLASS membership, SEX-differentiated", xlab = "Class", ylab = "Frequency", col = brewer.pal(3,"Set2"), beside = TRUE, legend = c("Female", "Infant", "Male"))
```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: (The distribution of Sex categories seems to have an appropriate distribution. Infants have a more significant number of cases in the A1 and A2 classes identified by their rings. It is interesting to find that some infant cases are observed in the A4 and A5 classes. These may be outliers and warrant further investigation. The SEX categories Female and Male have appropriate shapes closer to a normal distribution.)***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
work <- mydata[sample(1:nrow(mydata),200, replace = FALSE),]
plot(work[, 2:6])
```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
df <- as.data.frame(mydata)
ggplot(df, aes(VOLUME, WHOLE)) +
geom_point(aes(color = CLASS)) +
ggtitle(expression("Whole weight, in grams | Volume, cm"^3))
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
ratio <- max(df$SHUCK/ df$WHOLE)
ggplot(df, aes(WHOLE, SHUCK)) +
geom_point(aes(color = CLASS)) +
labs(title = " Shuck weight, in grams | Whole weight, in grams", x = "Whole weight", y = "Shuck weight") +
geom_abline(intercept = 0, slope = ratio)

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: (Initially, plot B has less variability than plot A, and the location of observations is more uniform. Class A5 (the older abalones) is generally congregated towards the bottom of the distribution. This observation suggests that as the abalone trees grow older, the shuck becomes a smaller percentage of the overall weight.)***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
par(mfrow = c(3,3))
hist(mydata[mydata$SEX == "F", "RATIO"], col = 6, main = "Female Ratio")
hist(mydata[mydata$SEX == "I", "RATIO"], col = 7, main = "Infant Ratio")
hist(mydata[mydata$SEX == "M", "RATIO"], col = 5, main = "Male Ratio")

boxplot(mydata[mydata$SEX == "F", "RATIO"], col = 6, main = "Female Ratio")
boxplot(mydata[mydata$SEX == "I", "RATIO"], col = 7, main = "Infant Ratio")
boxplot(mydata[mydata$SEX == "M", "RATIO"], col = 5, main = "Male Ratio")

qqnorm(mydata[mydata$SEX == "F", "RATIO"], col = 6, main = "Female Ratio")
qqline(mydata[mydata$SEX == "F", "RATIO"])

qqnorm(mydata[mydata$SEX == "I", "RATIO"], col = 7, main = "Infant Ratio")
qqline(mydata[mydata$SEX == "I", "RATIO"])

qqnorm(mydata[mydata$SEX == "M", "RATIO"], col = 5, main = "Male Ratio")
qqline(mydata[mydata$SEX == "M", "RATIO"])
```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: (Overall, the graphs closely represent normal distributions. However, all sex classes are slightly right-skewed. The right-skewed shape is evident from the shapes of the histograms and the deviations in the upper right on the normal line on all Q-Q plots. The box plots identify multiple outliers that contribute to the skewed distribution.)***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
f.ratio <- subset(mydata, mydata$RATIO %in% boxplot.stats(mydata[mydata$SEX == "F", "RATIO"])$out)
i.ratio <- subset(mydata, mydata$RATIO %in% boxplot.stats(mydata[mydata$SEX == "I", "RATIO"])$out)
m.ratio <- subset(mydata, mydata$RATIO %in% boxplot.stats(mydata[mydata$SEX == "M", "RATIO"])$out)

part <- rbind(f.ratio, i.ratio)
part_3b <- rbind(part, m.ratio)
kable(part_3b, format = "pipe")
```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: (There are 19 outliers in the data. There is one female outlier that has less rings than an infant. There is one female outlier that produces the largest ratio. Most of the outliers are infant.)***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
par(mfrow= c(2,2))
boxplot(VOLUME~CLASS, mydata, col = brewer.pal(5,"Set2"))
boxplot(WHOLE~CLASS, mydata, col = brewer.pal(5,"Set2"))
plot(mydata$RINGS, mydata$VOLUME, xlab = "RINGS", ylab = "VOLUME", col = brewer.pal(5,"Set2"))
plot(mydata$RINGS, mydata$WHOLE, xlab = "RINGS", ylab = "WHOLE", col = brewer.pal(5,"Set2"))
```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: (In this instance, weight and volume are not good predictors of age. The observations in the scatterplot indicate that a weak correlation between rings and volume/weight. However, by looking at the graphs we can tell that there is a correlation between volume and whole.)***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
agg.mean <- aggregate(x = mydata[c('VOLUME','SHUCK','RATIO')], by = list(mydata$SEX, mydata$CLASS), FUN=mean)

Volume.agg <- matrix(round(agg.mean$VOLUME, 2), nrow=3,dimnames = list(list('Female','Infant','Male'), list('A1','A2','A3','A4','A5')))

shuck.agg <- matrix(round(agg.mean$SHUCK, 2), nrow=3,dimnames = list(list('Female','Infant','Male'), list('A1','A2','A3','A4','A5')))

ratio.agg <- matrix(round(agg.mean$RATIO, 2), nrow=3,dimnames = list(list('Female','Infant','Male'), list('A1','A2','A3','A4','A5')))

print ("Volume")
Volume.agg
print ("Shuck")
shuck.agg
print ("Ratio")
ratio.agg
```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}

interaction.plot(mydata$CLASS, mydata$SEX, mydata$RATIO, fun = mean, xlab = "Class", ylab = "Ratio", main = "Mean Ratio per Class", col = brewer.pal(3, "Set2"), trace.label = "SEX", lty = 1, lwd = 2)

interaction.plot(mydata$CLASS, mydata$SEX, mydata$VOLUME, fun = mean, xlab = "Class", ylab = "Ratio", main = "Mean Ratio per Class", col = brewer.pal(3, "Set2"), trace.label = "SEX", lty = 1, lwd = 2)

interaction.plot(mydata$CLASS, mydata$SEX, mydata$SHUCK, fun = mean, xlab = "Class", ylab = "Ratio", main = "Mean Ratio per Class", col = brewer.pal(3, "Set2"), trace.label = "SEX", lty = 1, lwd = 2)
```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: (Enter your answer here.)***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
levels(mydata$SEX)[levels(mydata$SEX) =="F"] <- "adult"
levels(mydata$SEX)[levels(mydata$SEX) =="M"] <- "adult"
limit <- subset(mydata, RINGS < 16)
par(mfrow = c(2, 2))
boxplot(VOLUME~RINGS, subset(limit,limit$SEX == "I"),  col = 7, main = "Infant Volume | Rings", ylim = c(0,1100))
boxplot(VOLUME~RINGS, subset(limit,limit$SEX == "adult"),  col = 4, main = "Adult Volume | Rings", ylim = c(0,1100))
boxplot(WHOLE~RINGS, subset(limit,limit$SEX == "I"),  col = 7, main = "Infant Whole | Rings", ylim = c(0,400))
boxplot(WHOLE~RINGS, subset(limit,limit$SEX == "adult"),  col = 4, main = "Adult Whole | Rings", ylim = c(0,400))
```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: (The boxplots above illustrate a correlation between Volume and Whole weight. The tree growth has increased variability after ten rings. There is a larger range between abalones after ten rings. This increased degree of freedom helps explain the increased number of outliers amongst infants.)***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: (The main issue within this study was the delineation between SEX levels. It is unclear what constitutes physically between an infant and adult abalone. This mislabeling has led to much variability in the results based on weight and volume between the categories. There needs to be a more rigid classification for when an abalone can be identified as an adult. Better classification will give the data a better normal distribution, which will, in turn, support prediction.)***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: (I would need a firm understanding of the population statistics to accept the sample as representative. My line of questions would focus on the collection methods and information regarding the population. What is the population size? How was the data collected? When was the data collected? What was the length of time to collect the data? What is the population variance? What is the population distribution? )***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: ( Observational studies present difficulties based on the inability to determine causality. Within observational studies, researchers are unable to control the external effect on observances. Therefore, researchers are regulated to determine correlation because they cannot modulate variables to precisely determine causality. Observational studies do provide researchers the ability to find correlations for further analysis. Researchers can determine which variables they can control and observe to determine causality definitively.)***
