---
title: "COVID_FORECAST"
author: "THOMAS, JAMES"
date: "6/3/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, setup1, message = FALSE, warning = FALSE}
library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)
library(forecast)
library(xts)
```


```{r}
mydata <- read.csv("covid.csv", sep = ",", stringsAsFactors = TRUE)
```


```{r}
data <- data.frame(mydata)
```
##1)Descriptive Statistics: Do an Exploratory Data Analysis (EDA) and provide appropriate summary statistics / visualizations to help understand the spread of the disease (incidence) as well as its fatality rate. 

```{r}
str(data)

summary(data$case)

summary(data$deaths)


hist(data$cases)
skewness(data$cases)
kurtosis(data$cases)

hist(data$deaths)
skewness(data$deaths)
kurtosis(data$deaths)

case.rate <- data$popData2019 / data$cases
case.rate[is.na(case.rate)] <- 0
case.rate[!is.finite(case.rate)] <- 0
mean(case.rate)

fatality.rate <- data$deaths / data$cases
fatality.rate[is.na(fatality.rate)] <- 0
fatality.rate[!is.finite(fatality.rate)] <- 0

mean(fatality.rate)
sd(fatality.rate)


```

***#### Cases and fatalities are very low. The data is very skewed dut to many places an dates have zero cases or fatalities. ####***

##2)Inferential Statistics: Pick 2 countries and compare their incidence and fatality rates using hypothesis testing and confidence interval methods
```{r}
france <- subset(data, data$countriesAndTerritories == 'France')
poland <- subset(data, data$countriesAndTerritories == 'Poland')

#Summary France cases
summary(france$cases)
#Summary France Deaths  
summary(france$deaths)
#Summary Poland cases
summary(poland$cases)
#Summary Polancd Deaths   
summary(poland$cases)

#France mean deaths
france.mean <- mean(france$deaths)
france.mean

#France standard deviation of deaths
france.sd <- sd(france$deaths)
france.sd

#Poland mean deaths
poland.mean <- mean(poland$deaths)
poland.mean
#Poland standard deviation of deaths
poland.sd <- sd(poland$deaths)
poland.sd

t.test(france$cases, poland$cases)
#case differences are not statistically significant
t.test(france$deaths,poland$deaths)
#difference in deaths is statistically significant

france.daily.rate <- france$deaths / france$cases
france.daily.rate[is.na(france.daily.rate)] <- 0
france.daily.rate[!is.finite(france.daily.rate)] <- 0

poland.daily.rate <- poland$deaths / poland$cases

#Daily rates are statistically significant as well
t.test(france.daily.rate, poland.daily.rate)

```

##3)Correlation: Pick all the countries and evaluate the relationship between incidence rates and fatality rates. Compute the correlation coefficient, if relevant.
```{r}
cor(data$cases, data$deaths)
cor(fatality.rate, case.rate)
var.test(fatality.rate, case.rate)
```

***#### While cases and fatalities are correlated. Cases and fatalities rates have nearly no correlation. This is likely due to variance among death rates per country ####***

##4)Regression: Pick United States. Look at the time series of cases and time series of deaths. Use a regression model to predict the number of cases and the number of deaths for the next 5 days in the future.
```{r}
us <- subset(data, data$countriesAndTerritories =='United_States_of_America')

#Cases and deaths marginally correlated in US. a linear regression model may not produce the best results.
cor(us$cases, us$deaths)

us$dates <- as.Date(us$dateRep, format = '%d/%m/%y')

plot(us$dates, us$cases, col = 4, type = 'l', main = 'US COVID-19 Cases and Death', xlab = 'Date', ylab = 'Cases', lwd = 2)
lines(us$dates, us$deaths, col = 6, lwd = 2)
legend('topleft', legend=c("Cases", "Deaths"),
col=c(4,6), lty = 1, lwd = 2)


#Order the dates in the data frame to choose the appropriate days.
real.data <- us[order(us$dates),]

#data ends on 2020-12-31, list next 5 days
next5days <- as.Date(c("2020-12-31", "2021-01-01", "2021-01-02", "2021-01-03", "2021-01-04"))

days <- data.frame(dates = next5days)

#predict cases for next 5 days
model <- lm(cases ~ dates, data = real.data)
plot(model)

pred_1 <- data.frame(predict(model, newdata = days))
pred_1

#predict deaths for next 5 days
predcases <- as.numeric(c(123416.7,123820.9, 124225.0, 124629.1, 125033.3))
newcases <- data.frame(cases = predcases)
model2 <- lm(deaths ~ cases, data = real.data)
plot(model2)
pred_2 <- predict(model2, newdata = newcases)
pred_2
```

***#### Times series forecast to compare results ####***
```{r}
#create time series data to plot time series forecast
us.case.ts <- xts(us$cases, order.by = us$dates)
auto <- auto.arima(us.case.ts)
future <- forecast(auto, h=5)
plot(future, main = 'Forecast US Cases in Next 5 Days', xlab = 'Time', ylab = 'US Cases')

#Projections for cases over the next 5 Days
as.numeric(future$mean)

us.death.ts <- xts(us$deaths, order.by = us$dates)
auto2 <- auto.arima(us.death.ts)
future.deaths <- forecast(auto2, h = 5)
plot(future.deaths, main = 'Forecast Deaths in Next 5 Days', xlab = 'Time', ylab = 'US Deaths')

#Projections for deaths over the next 5 Days
as.numeric(future.deaths$mean)

```
